---
title: "Gene Set Enrichment Analysis"
author: "Joseph Descarreaux"
date: "`r Sys.Date()`"
output: html_document
---

First, we need to set our function to generate representation further in the analysis.

# Functions

First, the function used for the dotplot representations.
```{r fun_dotplot}
vis.dotplot <- function(obj, db, filename, path){

    # Make dotplot ----
    enrichplot::dotplot(
        object = obj,
        color = 'p.adjust',
        font.size = 10,
        title = paste(db, "Dotplot of enriched pathways", sep = ' - ')
    )

    # Save dotPlot
    ggsave(filename = filename, path = path )
}
```

Second, the function for the ridgeplot representation

```{r fun_ridgeplot}
vis.ridgeplot <- function(obj, db, filename, path){

    # Generate Ridge plot
    enrichplot::ridgeplot(
        x = obj,
        label_format = 30
    ) +
        # Reduce the character size for the axis informations
        theme(
            axis.text.x = element_text(size = 8),  # Adjust the x-axis text size
            axis.text.y = element_text(size = 8)   # Adjust the y-axis text size
        ) +
        labs(
            title = "Ridgeplot of enriched pathways",
            subtitle = paste('Database used : MSigDb', db, sep = ' - '),
            caption = Sys.time())

    # Save Ridge Plot
    ggsave(filename = filename, path = path )
}

```

Third. the function for the context network

```{r fun_cnetplo}
vis.cnetplot <- function(obj, ranked.list, db, path, filename){
    # Useful to visualize the relationship between enriched pathways
    enrichplot::cnetplot(
        obj,
        categorySize = "p.adjust",
        showCategory = nrow(obj@result),
        node_label = "category",
        foldChange = ranked.list,
        colorEdge = FALSE,
        cex_label_category = 0.5
    ) +
        # Personalize the color scale for the Fold Change
        scale_color_gradient2(name='log2(FC)', low='darkblue', high='firebrick') +
        # Add labels to the general plot
        labs(
            title = 'Graph representation of shared expressed gene',
            subtitle = paste('Database : MSigDB -', db, 'set'),
            caption = Sys.time())

    # Save Concept Network Plot
    ggsave(filename = filename, path = path)
}
```

Fourth, the homemade representation of the gseaplot with heatmap
```{r fun_gseaplot}
vis.gseaplot <- function(obj, path, txi.info){
    # Extract title information
    title <- obj@result$Description
    obj@result$Description <- ''
    # Extracted genes to subset the txi object for the heatmap
    enriched_genes <- obj@result$core_enrichment %>%
        stringr::str_split(., '/')
    # Generate the output directory
    out.gsea <- fs::path(path, 'plotGSEA')

    # Iterate over all enriched pathways
    for (i in seq_len(nrow(obj@result))) {
        # Generate GSEAplot
        gsea.plot <- enrichplot::gseaplot2(
            x = obj,
            geneSetID = i,
            color = "#86B875",
            title = paste(title[i],'(',txi.info$cell,'-',txi.info$hormone,')'),
            pvalue_table = TRUE,
            ES_geom = 'line'
        )

        # Susbet heatmap information
        heat.data <- txi$counts %>%
            as.data.frame %>%
            dplyr::select(matches(txi.info$cell)) %>%
            dplyr::select(matches(txi.info$hormone), matches('DMSO')) %>%
            dplyr::filter(rownames(.) %in% enriched_genes[[i]]) %>%
            setNames( sub('.*_(.*)_.*','\\1', colnames(.)))

        heat.plot <- pheatmap::pheatmap(
            heat.data,
            color = grDevices::colorRampPalette(
                rev(RColorBrewer::brewer.pal(n = 7, name ="RdBu")))(100),
            scale = 'row',
            cluster_cols = FALSE,
            treeheight_row = 0,
            cellwidth = 10,
            show_rownames = FALSE,
            na_col = 'darkgrey'
        )
        # Change obejct type from their respective type to ggplot
        gg.heat <- ggplotify::as.ggplot(heat.plot)
        gg.gsea <- ggplotify::as.ggplot(gsea.plot)

        # Concatenate plot unto a single pdf file
        gg.gsea + gg.heat + patchwork::plot_layout(widths = c(5,1))
        # Save plot
        ggsave(
            filename = paste(title[i] ,'gseaplot_heatmap.pdf', sep = '_'),
            path = out.gsea,
            height = 4.34, width = 8.31, units = 'in'
        )
    }
}
```

Additionnal, this is an little function to switch ENSEMBL gene ids, used in the
gsea analysis to their symbol counterpart, making it more readable for professionals
```{r fun_idSwitch}
excel.table <- function(df){
    df$core_enrichment <- df$core_enrichment %>%
        map(~ {
            paste(switchIds(stringr::str_split_1(string = .x, pattern = '/')),
                  collapse = '/')
        })
    df
}
```

# Running the GSEAnalysis

We first need to set our path accordingly. 
```{r path}
## Define paths ----
paths <- list(
    rds = './dgea_output/rds',
    output = './output_deseq2_pval0.01/gsea/clusterProfiler',
    kallisto = './deseq2_files/',
    anno = './files_util/anno'
)

paths$rds.output <- fs::path(paths$output, 'rds')
if(!dir.exists(paths$rds.output))dir.create(paths$rds.output, recursive = TRUE)

```
This is what each path should represent:

+ 'rds' : Directory to get the object generated from the DESeq2 analysis. 
+ 'output' : Directory to output the generated files
+ 'kalisto' : Directory to get the count data, used for the heatmap
+ 'anno' : Directory to the annotation file, used to generated the object of count data
+ 'rds.output' : Directory to store the ouput the object generated from the GSEA analysis

## Generate the count object

To get the count data, we need to generate a txi object from the kallisto data.

The following code will retrieve every needed file to load the count data.
```{r get_counts, eval=FALSE}
# Get kallisto files
k.files <- fs::dir_ls(paths$kallisto, regexp = '.h5$', recurse = TRUE) %>%
    '['(!grepl('DU145|ENZA',.))

# Get annotation informations
anno.file <- fs::dir_ls(paths$anno, regexp = 'protein_coding.csv$')

# Import full kallisto informations
txi <-  rnaseq::import_kallisto(
    filenames = k.files,
    anno = anno.file,
    ignoreTxVersion = TRUE
)

# Change column's name
samples <- fs::path_dir(k.files) %>% fs::path_file(.)
colnames(txi$counts) <-  samples
colnames(txi$abundance) <- samples
colnames(txi$length) <- samples
```
